---
title: "Analyzing Spatial Separation with Information Geometry"
author: "Phil Chodrow"
date: "January 18, 2017"
output: 
  html_document:
    theme: simplex
    highlight: tango
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This document is an introduction to information-geometric techniques for geospatial analysis. The core situation we consider is one in which our data consist of counts in two or more categories, distributed in space. A standard example is demographic census data, in which the categories might be racial groups, and the spatial units tracts. This document does not focus on mathematical details of the methods, which will be explained in a forthcoming paper. Rather, the focus is on the intuition behind various analytical methods, and how to carry them out in order to learn about spatial separation with freely-available software tools. The primary methods we use are implemented in the packages `compx` for the `R` programming language. Get `compx` by running: 

```{r, eval = FALSE}
# install.packages('devtools') if necessary
devtools::install_github('PhilChodrow/compx')
library(compx)
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(compx)
```

To prepare data and analyze outputs of `compx` functions, we will also need the following packages: 

```{r, warning = FALSE, message = FALSE}
library(tidyverse)
library(maptools)
library(igraph)
library(RColorBrewer) # for plots
```

# Data inputs

The functions provided by package `compx` assume that your data is expressed in two components. The first is a `SpatialPolygonsDataFrame` (`spdf`) containing geographic polygons ("tracts"). For illustrative purposes, `compx` comes bundled with an `spdf` with the Census tracts for Wayne County, Michigan, which includes the urban core of Detroit, an oft-analyzed city in quantitative segregation studies. The tracts were originally accessed via the `R` package `tigris`. 

```{r}
detroit_tracts %>% plot()
```

The second data input must be a demographic table with class in `c('tbl_df', 'tbl', 'data.frame')`, which I will refer colloquially to as a "data frame." `compx` includes an example in `detroit_race`, giving racial demographic counts for each tract for decennial Censuses 1970, 1980, 1990, 2000, 2010. Due to changing questions and collection methods, only 1990 data and later is comparable with 2010. The demographic data was assembled and normalized by the [Project on Diversity and Disparities](https://s4.ad.brown.edu/Projects/Diversity/Researcher/LTBDDload/Default.aspx) at Brown University.

```{r}
detroit_race
```

Note that `detroit_race` contains one row per combination of tract, time, and racial group. `compx` expects your demographic data to be in this tidy (or "long") format. If it's in a different shape, you may want to investigate the `dplyr` and `tidyr` packages to format it appropriately. 

Three columns are required: 

1. `tract`, the *key* relating the data the corresponding `spdf`. `compx` assumes that `tract` matches the `GEOID` column of `spdf@data`. 
2. `group`, the demographic category (such as racial group, in this case). 
3. `n`, the count of members of `group` in each `tract`. 

Additionally, you may include an optional column for time `t`. `compx` functions will automatically use this column of detected; if you don't want to do temporal analysis, you should delete the $t$ column if you have one. Any additional columns are ignored.  

## Data Preparation

Before going any further, let's do a bit of cleaning to ensure that we are only analyzing with tracts for which there is a reasonable amount of data available. 

```{r}
tracts_to_use <- detroit_race %>% 
	group_by(tract, t) %>%
	summarise(n = sum(n)) %>%
	filter(n > 100) 

tracts <- detroit_tracts[detroit_tracts@data$GEOID %in% tracts_to_use$tract, ]

f_tracts <- tracts %>%         # for plotting later
	fortify(region = 'GEOID')
```

# The Information Manifold

Our analytical approach is motivated by [*information geometry*](https://en.wikipedia.org/wiki/Information_geometry). The core notion of information geometry is to view the data as lying on an *information manifold*, and then use some simple tools of differential geometry to study that manifold's properties. 

The fundamental manifold property is the *metric tensor* $g$. The metric tensor contains complete information on how distances in information space are bent and curved, depending on where we are. It may therefore be used to compute distances. If $p, q, r \in M$, then, provided $p,q,$ and $r$ are sufficiently close, the *geodesic distance* between $q$ and $r$ is approximately 

$$ d(q, r) \approx \sqrt{g_p(q - r, q - r)}\;.$$

When $g = I$, the identity matrix, this formula reduces to $d(q,r) \approx \lVert q-r \rVert$, the Euclidean distance. When $g$ differs substantially from $I$, we get more interesting behavior. 

We can view the metric tensor $g$ as encoding the information we need to measure distances based on information, rather than on geography alone. A related and important idea of the metric tensor is that it encodes variation -- places where the components of the metric tensor are large correspond to boundary areas between different demographic clusters. 

## Computing the Metric Tensor

Because the metric tensor is fundamental to our approach, `compx` includes a function that computes it directly from `tracts` and appropriately keyed `data`. If the keyed data contains no column labeled `t`, then `compx` assumes that all data are from the same time period. The simplest case looks like this: 

```{r}
race_2010 <- detroit_race %>%
	filter(t == 2010) %>%
	select(-t)                 # remove the t column if not using it

metric_df <- compute_metric(tracts,             # use these tracts
							race_2010,          # and these data
							km = T,             # convert spatial units to km
							sigma = 100,        # for numerical derivatives
							divergence = 'euc') # use the Euclidean distance
```

In addition to the tracts and data parameters, `compute_metric` also allows users to specify whether length should be measured in km or degrees long/lat; the divergence function to use when comparing distributions, and `sigma`, a technical parameter relating to the computation of numerical derivatives required to compute the metric tensor. 

The output of `compute_metric` is a data frame, keyed by geoid, which gives the total population of the tract as well as a list-column with the metric tensor in local coordinates. Since we didn't give a time column, there are only two local coordinates *x* and *y*. 

```{r}
metric_df %>% head()
```

We can also take a quick peek at the metric tensor itself. For example, the tensors corresponding to the first two entries of `metric_df` are: 

```{r}
metric_df$local_g %>% head(2)
```

The metric tensor is a symmetric matrix. Roughly, $g_{ii}$ encodes the strength of the dependence on the frequency field on the $i$th coordinate. Thus, if we focus on 
```{r}
metric_df$local_g[[1]]
```
we can compare the entry of ```r metric_df$local_g[[1]][1,1] ``` in the `x` component to ```r metric_df$local_g[[1]][2,2] ``` for the `y` component to get a sense for the direction in which the demographic distribution is changing most rapidly. In this case, it's the $x$ direction. 

## Analyzing the Metric Tensor

To visualize the metric structure, it's usually necessary to apply a scalar function to the metric tensor. Good ones are the trace and the square root of the determinant, both of which are usefully interpretable from the information-geometric point of view. Roughly, the trace corresponds to length distortions, and the determinant to volume distortions in information space. 
```{r}
trace_df <- metric_df %>% 
	mutate(trace = map_dbl(local_g, . %>% diag() %>% sum()),
		   det   = map_dbl(local_g, . %>% det() %>% abs() %>% sqrt()))
trace_df %>% head()
```

It's useful to plot these scalars to see how they behave visually. Here's the trace. 

```{r, fig.height = 4}
f_tracts %>% 
	left_join(trace_df, by = c('id' = 'geoid')) %>% 
	ggplot() + 
	aes(x = long, y = lat, fill = trace, group = group) + 
	geom_polygon() + 
	viridis::scale_fill_viridis(option = 'magma', trans = 'log10') + 
	ggthemes::theme_map() 
```

One way to think about the more intense areas with higher traces is that they are areas of *transition*. If you compare this visualization to, e.g. UVA's [Racial Dot Map](https://demographics.virginia.edu/DotMap/index.html) you'll find that the lighter areas correspond to the predominating white/black boundary, as well as smaller subdivisions. 

## Incorporating Time

`compx` tries to make it easy for you to incorporate time into your analysis as well. To do this, you just need to make sure that your data has a `t` column with appropriate integer or numeric values. Then, `compute_metric` will work just as before. 

```{r}
race_temporal <- detroit_race %>%
	filter(t %in% c(1990, 2000, 2010))           # now we don't remove the t column
	
metric_df <- compute_metric(tracts, 
							race_temporal, 
							km = T, 
							sigma = 100, 
							divergence = 'euc')
```

This version of `metric_df` is different in two respects: 

```{r}
metric_df %>% head()
```

First, there's a `t` column for time. Second, the metric tensor is now `3 x 3`, which reflects the fact that we've added a third coordinate (time). 

```{r}
metric_df$local_g[[1]]
```

In general, it's not a good idea to directly compare temporal components of the metric tensor to spatial ones, since they have different units. In this case, the spatial components have units of kilometers, but the temporal ones have units of years. However, it's not wrong to interpret this result as saying that, at this point in space and time, the demographic composition changes more when you move a km in space than when you move a year in time. 


Just like before, we can apply some scalar functions to extract information about the metric tensor. The spatial trace is  the sum of the first two diagonal entries, corresponding to the components of the metric tensor that encode spatial (not temporal) variability. 

```{r, fig.height = 2.5, fig.width = 8}
trace_df <- metric_df %>% 
	mutate(trace = map_dbl(local_g, ~ .[1:2, 1:2] %>% diag() %>% sum()))

f_tracts %>% 
	left_join(trace_df, by = c('id' = 'geoid')) %>% 
	ggplot() + 
	aes(x = long, y = lat, fill = trace, group = group) + 
	geom_polygon() + 
	viridis::scale_fill_viridis(option = 'magma', trans = 'log10') + 
	ggthemes::theme_map() + 
	facet_wrap(~t)
```

The third component of the metric tensor quantifies the dependence on time. We can also extract that and visualize it: 

```{r, fig.height = 2.5, fig.width = 8}
t_df <- metric_df %>% 
	mutate(temporal = map_dbl(local_g, ~ .[3, 3]))

f_tracts %>% 
	left_join(t_df, by = c('id' = 'geoid')) %>% 
	ggplot() + 
	aes(x = long, y = lat, fill = temporal, group = group) + 
	geom_polygon() + 
	viridis::scale_fill_viridis(option = 'magma', trans = 'log10') + 
	ggthemes::theme_map() + 
	facet_wrap(~t)
```

We can see that, the predominantly white suburbs toward the west have generally experienced greater demographic change over time than the prodominantly black urban core toward the northeast. We also observe that the temporal component of the metric tensor tends to correlate with the spatial component of the previosu visualization. So, spatial transition areas also tend to be temporal transition areas. This makes sense if you think that the dynamics of population composition are continuous in space, so that change happens at the boundaries.  

```{r, fig.height = 2.5, fig.width = 8}

t_df  %>% 
	mutate(spatial = map_dbl(local_g, ~ .[1:2, 1:2] %>% diag() %>% sum())) %>% 
	ggplot() + 
	aes(x = spatial, y = temporal) + 
	geom_point() + 
	scale_y_continuous(trans = 'log') + 
	scale_x_continuous(trans = 'log') + 
	geom_smooth() + 
	facet_wrap(~t)

```


# Network Analysis

The other main form of analysis enabled by `compx` is network-based analysis used to quantify and identify structure in demographic variation. 

```{r}
g <- construct_information_graph(tracts, race_2010, divergence = 'DKL')
g
```

`g` is an `igraph` object with edge and node attributes. The `dist` edge attribute reflects how information-geometrically dissimilar are the connected nodes, where this dissimilarity is just the information distance between the nodes, calculated using the metric tensor. It's useful to visualize this network. In the plot below, dissimilar tracts have thin edges between them, while very similar ones are joined by thick edges. 

```{r, fig.height = 4}

edges <- g %>% as_long_data_frame() %>% tbl_df()
nodes <- data_frame(x = V(g)$x, y = V(g)$y)

ggplot() + 
    geom_polygon(aes(x = long, y = lat, group = group), fill = 'firebrick', alpha = .6, data = f_tracts) +
    geom_segment(aes(x = from_x, 
    				 y = from_y, 
    				 xend = to_x, 
    				 yend = to_y, 
    				 size = exp(-7 * dist^2)), color = 'black', data = edges) +
    ggthemes::theme_map() +
    scale_size_continuous(range = c(0,1)) + 
    guides(alpha = 'none', color = 'none', size = 'none')
```

Comparing to the figures above, you might notice that areas with thin or invisible edges correspond to the same "border" areas that we saw highlighted by the spatial trace above. 

## The Scale of Segregation
One useful way to analyze the spatial structure of segregation is via the eigenspectrum of the Laplacian matrix corresponding to the graph, a method inspired by [spectral clustering](http://www.cs.cmu.edu/~aarti/Class/10701/readings/Luxburg06_TR.pdf). In the plot below, large gaps in the spectrum correspond to strong "signals" in the spatial structure; intiutively, it "makes sense" to cut the structure into that many pieces. 
```{r, fig.height = 4}
A   <- affinity_matrix(g, sigma = 1)        # square exponential affinity with specified sigma
L   <- generalized_laplacian_matrix(A)      # L_{rw} in the tutorial cited above
evL <- eigen(L, symmetric = T)              # compute the spectrum

data.frame(n = 1:30, ev = 1 - rev(evL$values)[1:30]) %>%
	ggplot() +
	aes(x = n, y = ev) +
	geom_line() +
	geom_point() +
	scale_y_continuous(trans = 'log10') + 
	geom_vline(xintercept  = 2.5, linetype = 'dashed') + 
	geom_vline(xintercept  = 7.5, linetype = 'dashed') + 
	geom_vline(xintercept = 10.5, linetype = 'dashed') + 
	theme_bw()
```

For example, we see that in Detroit with under the KL divergence, there's a large gap after $n = 2$ and then a smaller one at $n = 7$. Of course, it's important to realize that there is some judgment required in identifying these hard cutoffs -- it's not a single cutoff but the spectrum of $A$ that most fully describes the community structure. 

## Spatial Community Detection

We can try to actually identify clusters based on the affinity matrix $A$. There are lots of approaches for finding clusters in an affinity matrix, but a simple and workable one is hierarchical clustering. 
```{r}
# do the clustering
clust <- hclust(as.dist(-A),  method = 'complete')

# assign node attributes to g giving the clusters
ncluster <- 7
g        <- g %>% set_vertex_attr(name = 'cluster', 
								  index = colnames(A), 
								  value = clust %>% cutree(ncluster))

# extract as data frame. We already extracted the edges df above. 
nodes    <- data_frame(x = V(g)$x, y = V(g)$y, cluster = V(g)$cluster)

ggplot() + 
    geom_segment(aes(x = from_x, 
    				 y = from_y, 
    				 xend = to_x, 
    				 yend = to_y, 
    				 size = exp(-7 * dist^2)), color = 'black', data = edges) +
	geom_point(aes(x = x, y = y, color = as.character(cluster)), size = 2, data = nodes) + 
    ggthemes::theme_map() +
    scale_size_continuous(range = c(0,1)) + 
    guides(alpha = 'none', color = 'none', size = 'none') + 
	scale_color_brewer(palette = 'Set1')
```

This is a reasonably strong result; the clustering has detected the predominating division between the white suburbs (green) and the predominantly black urban core (blue), as well as "Mexicantown" (red), the historically distinct community of Grosse Point (yellow), and Hamtramck (brown), an independent town with large European and Asian immigrant populations. Not all clusterings I've worked with have quite
such intuitive behavior, but all demonstrate promising performance. 

## Spatiotemporal Networks

It's also possible to use construct networks at multiple time-slices. Just like before, it's necessary to use data that has a `t` column with appropriate values. 

```{r}
g <- construct_information_graph(tracts, race_temporal, divergence = 'DKL')
```

When  `t` column is provided, the names of `g` are concatenations of the GEOID of the tract and the corresponding value of $t$. There's an edge between each tract and its corresponding step forward or backward in time. 

```{r}
g
```

We can also cluster on these temporal networks, in just the same fashion as we did before. 

```{r}
# construct the affinity matrix and do clustering
A     <- affinity_matrix(g, sigma = 1)
sigma <- 1
clust <- hclust(as.dist(-A), method = 'complete')

# assign clusters as a node attribute
ncluster <- 12
g        <- g %>% set_vertex_attr(name = 'cluster', 
								  index = colnames(A), 
								  value = clust %>% cutree(ncluster))
```

It's more complex to visualize these networks in a workable way, but not hard to visualize the clusters over time: 

```{r, fig.height = 2.5, fig.width = 8}

# get the edges, only including ones that are "within" a time slice. 
edges <- g %>% 
	as_long_data_frame() %>% 
	tbl_df() %>% 
	mutate(type = ifelse(from_t == to_t, 'spatial', 'temporal')) %>% 
	filter(type == 'spatial') %>% 
	mutate(t = as.integer(stringr::str_sub(from_t, -5)))

# get the nodes, including temporal and cluster info
nodes <- data_frame(x = V(g)$x, 
					y = V(g)$y, 
					t = V(g)$t, 
					cluster = V(g)$cluster)

# plot the clusters, faceting on time
pal   <- brewer.pal(9, name = "Set1") %>%
	colorRampPalette()

ggplot() + 
	geom_segment(aes(x = from_x, 
					 y = from_y, 
					 xend = to_x, 
					 yend = to_y, 
					 size = exp(-7 * dist^2)), data = edges, color = 'grey40') + 
	geom_point(aes(x = x, y = y, color = as.character(cluster)), size = 1, data = nodes) + 
	facet_wrap(~t) + 
	scale_size_continuous(range = c(0,1)) + 
	ggthemes::theme_map() + 
	guides(color = FALSE, size = FALSE) + 
	scale_color_manual(values = pal(ncluster)) 
```

Currently, this temporal clustering method has some limitations -- tracts don't change demographics very quickly over the time period we are considering, and so clusters tend to be quite persistent. This in itself isn't a bad thing per se, but it's also the case that we generate some slightly counterintuitive clusters this way. Getting useful clusterings is a major area of focus ahead. 

However, there's also a substantial amount of signal in the clusters. For example, the clustering detects that the historically white community of Grosse Point (the far eastern tip) has somewhat receded from 1990 to 2010, with its western-most areas being "absorbed" into the predominantly black cluster. This makes sense when we visualize how relative percentages of White and Black residents have changed over time:  

```{r}

# construct a data frame of percentages. 
percent_df <- race_temporal %>% 
	filter(t %in% c(1990, 2010)) %>% 
	group_by(tract, t) %>% 
	mutate(percent = n /sum(n)) %>%
	rename(race = group) %>% 
	filter(race %in% c('Black', 'White')) 

# plot the df 
pal <- brewer.pal(9, 'Blues') %>% 
	colorRampPalette()

f_tracts %>%
	left_join(percent_df, by = c('id' = 'tract')) %>% 
	tbl_df() %>% 
	ggplot() +
	aes(x = long, y = lat, group = group, fill = percent) + 
	geom_polygon() + 
	facet_grid(t ~ race) + 
	ggthemes::theme_map() + 
	scale_fill_distiller(palette = 'BuPu', direction = 1) + 
	theme(panel.background = element_rect(fill = 'grey80'))
	




```

# Future Work

Priorities for future work on `compx` include refining clustering methods, performance enhancements, and improvements TBD based on suggestions from those in the planning and sociological communities. 

